A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-24 03-semantics.tex
Renomeia apêndice
Adiciona parágrafo sobre memory bound, Seção 3.3.3

The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions which only
apply to the program as a whole, and are the only to manipulate the stack
level:
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[\JOT]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$\ceu{p=\Nop}\vee\isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

We now state and prove the main result of this section,
Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
The idea of the proof is that a series of~$\nst$ or~$\out$ transitions
eventually decrease the rank of the current description until an irreducible
description is reached.  This irreducible description is the final result of
the reaction.
%%
\begin{theorem}[Termination]\label{thm.x.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.x.rank-nst-*},
    \[
      \<i,j>=\<0,0>\ge\rank(\delta')\,,
    \]
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases depending on whether or not $\delta$~is
    nested-irreducible.
    \begin{case}
      \label{thm.x.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
      that $j>0$.  Then there are two subcases.
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nonnil}
        $e\ne\nil$.
        %%
        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
        are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
        $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
        Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
        \begin{align*}
          \rank(\delta)=\rank(\delta_1')=\<i,j>
                                        \ge\rank(\delta')=\<i',j'>\,.
        \end{align*}
        If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
        similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
        was consumed by the nested transitions.  Thus,
        $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
        $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
        $\delta\trans[*]\delta''_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nil}
        $e=\nil$.
        %%
        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta''$.  By
        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
        Therefore, $\delta\trans[*]\delta''_\#$.
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      Then~$e=\nil$ and, by Theorems~\ref{thm.x.rank-nst-*}
      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
      that~$\delta\nst[*]\delta'_\Hnst$
      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
    \end{case}
  \end{induction}
\end{proof}

\subsubsection*{Memory bounds}

TODO: Acho que sai direto do~$\pot(p)$.

- program is finite
- lexical scope
    - no heap allocation
- no code reentrancy
    - reexecution only due to loops
    - loop reuse nested vars
-

>>>>

The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions which only
apply to the program as a whole, and are the only to manipulate the stack
level:
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[\JOT]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$p=\ceu{\Nop,\Break}\vee\isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

We now state and prove the main result of this section,
Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
The idea of the proof is that a series of~$\nst$ or~$\out$ transitions
eventually decrease the rank of the current description until an irreducible
description is reached.  This irreducible description is the final result of
the reaction.
%%
\begin{theorem}[Termination]\label{thm.x.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.x.rank-nst-*},
    $\<i,j>=\<0,0>\ge\rank(\delta')$,
    %%
    % \[
    %   \<i,j>=\<0,0>\ge\rank(\delta')\,,
    % \]
    %%
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases depending on whether or not $\delta$~is
    nested-irreducible.
    \begin{case}
      \label{thm.x.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
      that $j>0$.  Then there are two subcases.
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nonnil}
        $e\ne\nil$.
        %%
        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
        are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
        $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
        Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
        \begin{align*}
          \rank(\delta)=\rank(\delta_1')=\<i,j>
                                        \ge\rank(\delta')=\<i',j'>\,.
        \end{align*}
        If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
        similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
        was consumed by the nested transitions.  Thus,
        $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
        $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
        $\delta\trans[*]\delta''_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nil}
        $e=\nil$.
        %%
        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta''$.  By
        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
        Therefore, $\delta\trans[*]\delta''_\#$.
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      Then~$e=\nil$ and, by Theorems~\ref{thm.x.rank-nst-*}
      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
      that~$\delta\nst[*]\delta'_\Hnst$
      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
    \end{case}
  \end{induction}
\end{proof}

\subsubsection{Memory bound}

As \CEU has no mechanism for heap allocation, unbounded iteration, or
general recursion, the maximum memory usage of a given \CEU program is
determined solely by the length of its code, the number of variables it
uses, and the size of the event stack that it requires to run.  The code
length and the number of variables used are easily determined by code
inspection.  The maximum size of the event stack during a reaction of
program~$p$ to external event~$e$ corresponds to~$\pot(p,e)$, i.e., to the
maximum number of internal events that~$p$ may emit in a reaction to~$e$.
If~$p$ may react to external events~$e_1$, \dots,~$e_n$ then, in the worst
case, its event stack will need to
store~$\max\{\pot(p,e_1),\dots,\pot(p,e_n)\}$ events.

% - program is finite
% - lexical scope
%     - no heap allocation
% - no code reentrancy
%     - reexecution only due to loops
%     - loop reuse nested vars
